---
title: "Event valence"
subtitle: "Validation of the Bayesian ordinal mixed model"
author: "Szymon Talaga"
date: last-modified
date-format: "MMMM D, YYYY | HH:mm:ss (Z)"
toc: true
toc-depth: 3
toc-location: left
execute:
  echo: false
  warning: false
  cache: false
  daemon: false
format:
  html:
    grid:
      margin-width: 350px
    embed-resources: true
  pdf:
    toc: false
    number-sections: true
    colorlinks: true
reference-location: margin
citation-location: margin
---

```{r}
library(reticulate)
library(stringr)
library(arrow)
library(dplyr)
library(tibble)
library(brms)
library(emmeans)

set.seed(303L)

use_python(normalizePath(R.home("../../bin/python")), required = TRUE)

project <- import("project")
config  <- project$config
paths   <- project$paths
here    <- paths$analyses / "valence"

countries <- config$countries$order
countries <- countries[0L:length(countries)]
```

```{r}
glmm   <- readRDS(as.character(paths$glmm / "valence" / "event.rds"))
data   <- tibble(glmm$data) %>% select(event:day)
sample <- sample_n(data, 1e5L)
```


# Posterior Predictive Checks

```{r}
ppd <- posterior_predict(glmm, newdata = sample, cores = 8L)
```

```{r}
ppc_bars <- pp_check(glmm, newdata = sample, type = "bars", cores = 8L)
```


```{r}
bayesplot::pp_check(glmm, type = "stat", stat = "mean")
```

```{r}
# Predictive check by categories
bayesplot::pp_check(glmm, type = "stat_grouped", stat = "mean", group = "country")
```

```{r}
# Distribution-based check
bayesplot::pp_check(glmm, type = "dens_overlay", ndraws = 100)
```

# Posterior Distribution of Fixed Effects

## Fixed Effects Point Estimates and Credible Intervals

```{r}
fe_samples <- as_draws_df(glmm, variable = "b_", regex = TRUE)

cat("Fixed effects samples dimensions:", dim(fe_samples), "\n")
cat("(posterior draws, parameters)\n")
```

```{r}
# Extract and display fixed effects with credible intervals
fe_est <- fixef(glmm)
cat("\nFixed Effects Estimates (95% CrI):\n")
print(fe_est)
```

## Posterior Predictive Distributions by Category

Generate predictions for representative combinations of predictors:

```{r}
library(tidybayes)

# Create prediction data: all combinations of country and political
pred_data <- expand_grid(
    country = unique(data$country),
    political = levels(data$political),
    event = NA  # outcome to predict
)

# Get posterior epred (expected value on response scale)
epred <- posterior_epred(glmm, newdata = pred_data, re_formula = NA)

cat("Posterior expected value predictions shape:", dim(epred), "\n")
cat("(posterior draws, prediction scenarios)\n")
```

```{r}
# Summarize posterior predictions
pred_summary <- pred_data %>%
    mutate(
        .epred_mean = colMeans(epred),
        .epred_q05 = apply(epred, 2, quantile, probs = 0.05),
        .epred_q95 = apply(epred, 2, quantile, probs = 0.95),
    ) %>%
    arrange(country, political)

cat("\nPosterior Predicted Expected Values (95% CrI):\n")
print(pred_summary, n = nrow(pred_summary))
```

# Leave-One-Out Cross-Validation (LOO-IC)

## LOO-IC Diagnostics

LOO-IC provides model comparison and out-of-sample predictive accuracy assessment:

```{r}
# Compute LOO-IC (this may take a moment)
loo_fit <- loo(glmm)

# Display LOO diagnostics
cat("\nLOO Information Criterion:\n")
print(loo_fit)
```

## Pareto k Diagnostics

Pareto k values diagnose the reliability of LOO-IC. Values > 0.7 are problematic:

```{r}
# Extract Pareto k values
k_vals <- loo_fit$diagnostics$pareto_k

cat("\nPareto k Diagnostic Summary:\n")
cat("  Min k:", round(min(k_vals), 3), "\n")
cat("  Max k:", round(max(k_vals), 3), "\n")
cat("  Mean k:", round(mean(k_vals), 3), "\n")
cat("  Observations with k > 0.7 (highly influential):", sum(k_vals > 0.7), "\n")
cat("  Observations with k > 0.5 (moderately influential):", sum(k_vals > 0.5), "\n")

if (sum(k_vals > 0.7) > 0) {
    cat("\nWarning: Some observations are highly influential for cross-validation.\n")
    cat("This may indicate model misspecification or outliers.\n")
}
```

```{python}
# Visualize Pareto k diagnostic
import numpy as np
import matplotlib.pyplot as plt
import seaborn as sns

k_vals = np.array(r.k_vals)

fig, ax = plt.subplots(figsize=(10, 4))
ax.scatter(range(len(k_vals)), sorted(k_vals), alpha=0.6, s=20)
ax.axhline(0.5, color="orange", linestyle="--", alpha=0.7, label="Moderately influential (k > 0.5)")
ax.axhline(0.7, color="red", linestyle="--", alpha=0.7, label="Highly influential (k > 0.7)")
ax.set_xlabel("Observation (sorted by k)")
ax.set_ylabel("Pareto k")
ax.legend()
ax.set_title("Pareto k Diagnostics for LOO-IC")
ax.grid(True, alpha=0.3)
fig.tight_layout()
fig.savefig(as.character(r.here / "pareto-k-diagnostic.png"), dpi=100, bbox_inches="tight")
plt.close()
```

# Model Predictive Performance

## Prediction Accuracy for Ordinal Outcomes

```{r}
library(tidybayes)

# Get posterior predictions for each observation
post_pred <- posterior_predict(glmm, ndraws = 500)
pred_mode <- apply(post_pred, 2, function(x) as.numeric(names(table(x)))[which.max(table(x))])

# Compare with observed
comparison <- tibble(
    observed = as.integer(data$event),
    predicted = pred_mode
) %>%
    mutate(
        correct = observed == predicted,
        abs_error = abs(observed - predicted)
    )

cat("\nPredictive Accuracy for Event Valence:\n")
cat("  Overall accuracy:", round(mean(comparison$correct), 3), "\n")
cat("  Mean absolute error:", round(mean(comparison$abs_error), 3), "\n")

# Accuracy by category
cat("\nAccuracy by Political Classification:\n")
acc_political <- data %>%
    mutate(
        observed = as.integer(event),
        predicted = pred_mode,
        correct = observed == predicted
    ) %>%
    group_by(political) %>%
    summarize(
        n = n(),
        accuracy = mean(correct),
        mae = mean(abs(observed - predicted))
    )
print(acc_political)
```

## Confusion Matrix (Ordinal Scale)

```{r}
library(caret)

# Create confusion matrix
conf_matrix <- confusionMatrix(
    factor(comparison$predicted, levels = 1:nlevels(data$event)),
    factor(comparison$observed, levels = 1:nlevels(data$event))
)

cat("\nConfusion Matrix:\n")
print(conf_matrix$table)

cat("\nPer-Class Metrics:\n")
print(conf_matrix$byClass)
```

# Summary Assessment

```{r}
cat("\n=== MODEL VALIDATION SUMMARY ===\n\n")

cat("1. CONVERGENCE: ",
    if (max(rhat_vals, na.rm = TRUE) <= 1.01) "✓ Good" else "✗ Check Rhat values", "\n")

cat("2. ESS (Effective Sample Size): ",
    if (low_bulk == 0 & low_tail == 0) "✓ Adequate" else "⚠ Some low ESS values", "\n")

cat("3. POSTERIOR PREDICTIVE CHECK: Visual inspection recommended\n")

cat("4. PREDICTION ACCURACY: ",
    round(mean(comparison$correct), 1), "% correct\n")

cat("5. LOO-IC RELIABILITY: ",
    if (sum(k_vals > 0.7) == 0) "✓ Reliable" else paste("⚠", sum(k_vals > 0.7), "problematic observations"), "\n")

cat("\n==============================\n")
```
